---
title: "Faraway Ch3 Counting Regression"
author: "Cora Allen-Coleman"
date: "1/17/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(faraway)
```

read and clean data
```{r}
data(gala)
gala <- gala[,-2] #remove endemics variable, not used in this analysis
```

Linear Regression Predicting Number of Species on an Island
```{r linear regression}
modl <- lm(Species ~ . , gala)
plot(predict(modl),residuals(modl),xlab="Fitted",ylab="Residuals") #nonconstant variance
```

Linear Regression Predicting sqrt(Number of Species on an Island)
```{r transformed}
modt <- lm(sqrt(Species) ~ . , gala)
plot(predict(modt),residuals(modt),xlab="Fitted",ylab="Residuals") #better variance
summary(modt)
```

Poisson Model
```{r poisson}
modp <- glm(Species ~ .,family=poisson, gala)
summary(modp)
```

Check Poisson Model outliers
```{r}
halfnorm(residuals(modp)) #check if large deviance can be explained by an outlier
```

Check Poisson Assumptions
For a Poisson distribution, mean is equal to variance.
To estimate the variance for each value of mean, use (y-u_hat)^2
```{r}
plot(log(fitted(modp)),log((gala$Species-fitted(modp))^2), #fit (mean) vs logvariance
  xlab=expression(hat(mu)),
  ylab=expression((y-hat(mu))^2))
abline(0,1) #line representing mean equal to variance (run in console)
```
Variance is generally larger than the mean. "When the variance assumption of the Poisson regression model is broken but the link function and choice of predictors is correct, the estimates of β are consistent, but the standard errors will be wrong. We cannot determine which predictors are statistically significant in the above model using the output we have." This is overdispersed. Add a dispersion parameter to fix this.

After adding a dispersion parameter $\phi$, $\lambda$ is distributed as a gamma random variance with mean $\mu$ and variance $\frac{\mu}{\phi}$. The outcome y is distributed as negative binomial with mean = $\mu$ and variance = $y = \frac{\mu(1+\phi)}{\phi}$. This accounts for the overdispersion we see. We can estimate this dispersion parameter.

Accounting for Overdispersion
```{r overdispersion}
(dp <-sum(residuals(modp,type="pearson")^2)/modp$df.res) #estimate phi
summary (modp, dispersion=dp)
```
"the estimation of the dispersion and the regression parameters is independent" To compare Poisson models with overdispersion, use F test rather than chi square test.

Rate Models
modeling ratios
```{r read in data}
data(dicentric)
round(xtabs(ca/cells ~ doseamt+doserate, dicentric),2)
with(dicentric,interaction.plot(doseamt,doserate,ca/cells))
```
abnormalities/total cells

Linear Model for Rate Outcome
```{r}
lmod <- lm(ca/cells ~ log(doserate)*factor(doseamt), dicentric)
summary(lmod)
#diagnostics
plot(residuals(lmod) ~ fitted(lmod),xlab="Fitted",ylab="Residuals")
abline(h=0) #run in console
```
Diagnostic plot shows residuals increasing as fitted values increase.

Directly Model Count Response
```{r}
dicentric$dosef <- factor(dicentric$doseamt)
pmod <- glm(ca ~ log(cells)+log(doserate)*dosef, family=poisson, dicentric)
summary(pmod)
```
this uses log(cells) as a predictor. Because the coefficient for log(cells) is close to 1, we could keep this constant by using it as an offset

Rate Model with Offset
```{r offset model}
rmod <- glm(ca ~ offset(log(cells))+log(doserate)*dosef, family=poisson,dicentric)
summary(rmod)
```

Negative Binomial
"arises from the generalization of the Poisson where the parameter λ is gamma distributed."
```{r poisson}
data(solder)
soldmodp <- glm(skips ~ . , family=poisson, data=solder)
deviance(soldmodp)
df.residual(soldmodp)

#adding interactions
modp2 <- glm(skips ~ (Opening +Solder + Mask + PadType + Panel)^2, family=poisson, data=solder)
deviance(modp2) #deviance still too large
pchisq(deviance(modp2),df.residual(modp2),lower=FALSE)
```

```{r neg binomial model}
library(MASS)
modnb <- glm(skips ~ . , negative.binomial(1),solder); modnb
#allow k to vary
modnb <- glm.nb(skips ~ .,solder)
summary(modnb)
```